\chapter{First-set Algorithm}
\label{chapter:first-set}

In \Cref{chapter:wf-algorithm},
we presented the well-formedness check
implemented in \lpeg{},
which ensures that the input PEG is complete.
However, this is not the only role of this algorithm:
It also ensures that other algorithms
implemented in \lpeg{} terminate,
as they traverse patterns in similar ways.
This chapter presents one such algorithm,
which we label as the first-set algorithm.

The first-set algorithm is responsible for
computing the set of first characters that
can possibly be accepted by a pattern,
and a Boolean value
that indicates whether the pattern may
accept the empty string.
We call this Boolean value the
\emph{emptiness} value of the pattern.
The key properties of the algorithm,
which we later prove,
are the following:
Any pattern fails any string that starts
with a character that is not in the first-set
of the pattern;
and the pattern fails the empty string
if its emptiness value is $false$.

Both are conservative approximations,
which means that a full first-set
and an emptiness value of $true$ are
the safest options, yet the least useful ones.
That is because \lpeg{} uses this algorithm and its properties
to optimize certain patterns in the code generation phase.
So, ideally, we would like the first-set to be as small as possible,
and the emptiness value of $false$ as common as possible,
so that \lpeg{} is able to optimize code more often.

One of the patterns that \lpeg{} tries to optimize
using this algorithm is the ordered choice.
As a base reference, we display below the code
that \lpeg{} generates for an ordered choice $\PChoice{p_1}{p_2}$
without any optimizations.
\begin{align*}
    & Choice(L_2) \\
    & p_1 \\
    & Commit(L_{end}) \\
    L_2:\ & p_2 \\
    L_{end}:\ &
\end{align*}

This unoptimized code starts with a choice instruction,
which creates a checkpoint for the initial match state,
so that, if $p_1$ fails,
it is able to restore this state
before running $p_2$.
If, instead, $p_1$ succeeds, it runs a commit instruction,
which deletes this checkpoint and jumps to $L_{end}$.

However,
if $p_1$ and $p_2$ have disjoint first-sets,
and $p_1$ has an emptiness value of $false$,
then \lpeg{} generates the following optimized code.
\begin{align*}
    & TestSet(first(p_1), L_2) \\
    & p_1 \\
    & Jump (L_{end}) \\
    L_2:\ & p_2 \\
    L_{end}:\ &
\end{align*}

This optimized code begins with an instruction (test set)
that checks whether the input string starts with
a character in the first-set of $p_1$,
and jumps to $L_2$ if not.
In the case where the input string is empty,
we know that $p_1$ would fail,
because $p_1$ has an emptiness value of $false$.
If, otherwise, the input string starts with a character
that is not in the first-set of $p_1$,
then we know that $p_1$ would also fail,
by the key property of first-sets.
So, in either case, we know that $p_1$ would fail,
making the choice pattern equivalent to $p_2$.
That is why, in the case of failure,
the test-set instruction jumps to the code of $p_2$,
without ever running the code of $p_1$.

Meanwhile, the test-set instruction succeeds
if the input string starts with a character
that \emph{is} in the first-set of $p_1$,
and, therefore, is non-empty.
In this case, because the first-sets of $p_1$ and $p_2$ are disjoint,
we know that this character is \emph{not} in the first-set of $p_2$.
And, from the key property of first-sets,
this means that $p_2$ would fail,
making the choice pattern equivalent to $p_1$.
So, in this case, \lpeg{} simply executes the code of $p_1$.

This optimization significantly improves
the performance \lpeg{} when matching some choice patterns.
This improvement is possible
by replacing the costly creation/restoration/deletion of checkpoints
with inexpensive jumps and inspections on the input string.

In this chapter, we discuss the first-set algorithm
and prove its key properties.

\input{chapters/optimizations/intuition}
\input{chapters/optimizations/emptymatchable}
\input{chapters/optimizations/definition}
\input{chapters/optimizations/optimization}